Analyzing Student Use of Spacing and Interleaving Strategies in Interactions with GenAI-Powered Chatbots in Programming Courses

### 2024-11-13 CSEDU2025/Paper.tex
update paper

\usepackage{epsfig}
\usepackage{subcaption}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{multicol}
\usepackage{pslatex}
\usepackage{algorithm2e}
\usepackage[bottom]{footmisc}
% Please add other packages that you may need BEFORE the SCITEPRESS.sty package.
\usepackage{natbib}
\usepackage{multirow}
\usepackage{SCITEPRESS}
\usepackage{arydshln}

@@

\cite{Margulieux24} conducted a study on how undergraduate students in
introductory programming courses used generative AI to solve programming
problems in a naturalistic setting. The research focused on examining the
relationship between AI usage and students’ self-regulation strategies,
self-efficacy, and fear of failure in programming. Furthermore, the study
explored how these variables interacted with the characteristics of the learners,
the perceived usefulness of AI, and academic performance. The findings revealed
that students with higher self-efficacy, lower fear of failure or higher prior
grades tended to use AI less frequently or later in the problem solving process
and perceived it as less useful compared to their peers. However, no significant
relationship was found between students’ self-regulation strategies and their
use of AI.

\cite{Bastani24} investigated the impact of generative artificial intelligence
(AI), specifically GPT-4, on human learning, with a focus on mathematics
education in a high school. The problem addressed was how the use of generative
AI could affect the acquisition of new skills, which is crucial for long-term
productivity. The method involved a controlled experiment with approximately
one thousand students, who were exposed to two GPT-4-based tutors: a simple
tutor (GPT Base) and another with safeguards designed to promote learning
(GPT Tutor). The results showed that while access to GPT-4 improved performance
on practice exercises (48\% with GPT Base and 127\% with GPT Tutor), the removal
of access to GPT Base led to a 17\% decrease in student performance on exams.
This indicated that unrestricted use of *GPT Base* could hinder learning.
However, the GPT Tutor was able to mitigate this negative effect.

\cite{Boudouaia24} investigated the impact of ChatGPT-facilitated programming on
college students’ programming behaviors, performance, and perceptions. The
problem addressed was whether the use of ChatGPT in programming education could
influence how students interacted with programming tasks, their learning
outcomes, and their attitudes toward the technology. The study employed a
quasi-experimental design involving 82 college students divided into two
groups: one using ChatGPT-assisted programming (CFP) and the other using
traditional self-directed programming (SDP). The method combined mixed data
collection techniques, including behavioral logs, programming performance
evaluations, and interviews. The results showed that students using the CFP mode
exhibited more frequent behaviors such as debugging and reading feedback.
Although CFP students had a higher average score in programming performance,
there was no statistically significant difference compared to the SDP group
However, students’ perceptions of ChatGPT improved significantly after the
intervention, with increased perceived usefulness, ease of use, and intention to
continue using the tool in the future.

\cite{Sidra24} addressed the growing importance of metacognitive skills in an
AI-enabled workforce. The problem discussed was how AI systems, especially
collaborative tools like large language models (LLMs), impacted human
decision-making and the challenges they posed due to the lack of human-like
metacognition in AI. The method used identified four main characteristics that
differentiated human-AI interaction from human-human collaboration and proposed,
based on the dual-process theory of human thought, that metacognitive thinking
was crucial for the effective use of AI. The results indicated that human
workers needed to develop and apply metacognitive skills, such as planning,
monitoring, and evaluating their thought processes, to mitigate cognitive biases
and improve decision-making when working with AI. The study concluded by
recommending training and AI design improvements to enhance workers’
metacognitive abilities, ensuring more effective collaboration with AI systems.

Learning based on metacognitive strategies has been shown to be effective in
enhancing students’ performance and skills. \cite{Zheng19} investigated the
effects of metacognitive scaffolds on group behavior, performance, and cognitive
load in computer-supported collaborative environments. The results showed
positive impacts on metacognitive behavior and group performance without
increasing the cognitive load. \cite{LiWei23} proposed a metacognition-based
collaborative approach to enhance performance in collaborative programming. The
findings indicated that this approach influenced computational thinking,
critical thinking, and metacognitive awareness. Similarly, \cite{Wang23}
explored how metacognitive instruction affects computational thinking, critical
thinking, and metacognitive skills in collaborative programming, observing
improvements in problem-solving and collaboration. \cite{Khusnul24} investigates
the efficacy of a meta-learning approach in improving metacognitive and creative
skills. This study confirms the effectiveness of meta-learning strategies and
elucidates the relationship between meta-learning and metacognition. This
research suggests that metalearning can improve metacognitive abilities,
providing valuable insights into educational technology and course design in
higher education settings.

@@

CharlieBot is an educational chatbot built on ChatGPT 3.5 and enhanced with
Retrieval-Augmented Generation (RAG) technique to improve the precision of
responses provided to students \cite{Sun24}. The RAG approach was developed
using materials from a MOOC, including text, exercises, video transcripts, and
other resources.

@@

The study was conducted in three phases: data collection,
categorization, and analysis. During the data collection phase, students
enrolled in a second semester Java course at a public university in Spain were
introduced to CharlieBot and allowed to use it freely, without being subjected
to a specific methodology for its use. Data were collected anonymously, ensuring
that the records of interactions could not be linked to any individual student
or their academic performance.

In the categorization phase, the students' messages were manually classified
into eight distinct categories for analysis, representing different types of
prompt that students use when interacting with the chatbot. Initially,
\cite{Ghimire24} proposed four categories: Debugging Help (DH), Conceptual
Question (CQ), Code Snippet (CS), and Complete Solution (CSO). However, the data
collected indicated the need for additional categories, leading to the inclusion
of four more: Multiple Question (MQ), Student Corrections (SC), Language Change
(LC), and Uncategorized (U). In this study, students were free to use the
chatbot as they wanted, which contributed to the emergence of these additional
categories. Table \ref{tab:categories} presents these categories along with
their respective descriptions. To facilitate the analysis of interleaving, the
prompt topics were also categorized according to the course subjects, which are:
(1) Java, (2) Object Orientation, (3) Testing, (4) Recursion, (5) Data
Structures, and (6) Sorting and Searching Algorithms.

@@

A total of 625 student messages were categorized in 81 conversations, with
an average of 7.72 messages per conversation. This data offers insights into how
users engage with CharlieBot and the types of prompts they submit. The results
of the analysis are presented in this section, addressing the research questions
outlined in the introduction. Figure \ref{fig:graph1} shows the distribution of
the messages in the categories. The following sections present the results and
discussions of each research questions.

@@

Most of the messages are located in categories Conceptual Question and Debugging
Help corroborate the findings of \cite{Ghimire24} which showed that the
questions are also localized in the same categories.

Multiple Question exercise resolutions represented 14.9\% of the responses.
These prompts typically involved students submitting one or more exercises to
CharlieBot and requesting solutions. The conversation 13 in \ref{fig:graph2}
illustrates a sequence of multiple choice questions.

@@

Figure \ref{fig:graph1} also indicates that approximately 5.8\% of the prompts
written by students were classified as Uncategorized. These messages include
expressions of gratitude toward the chatbot, such as a simple '\textit{thanks}',
contextual statements like '\textit{I'm reviewing object orientation}', as well
as requests for additional exercises and summaries. Depending on the type of
investigation, there is opportunity for these messages to be further subdivided
into new categories.

@@

Despite some occasional variation (3.54), prompts classified as Uncategorized
exhibit a relatively low average time (1.65) for students to resume interaction
with the chatbot. This variation occurs, in part, due to situations where students
end the conversation with an uncategorized message, such as a simple
\textit{'thanks'} but later choose to resume the interaction.

>>>>

\usepackage{epsfig}
\usepackage{subcaption}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{multicol}
\usepackage{pslatex}
\usepackage{algorithm2e}
\usepackage[bottom]{footmisc}
% Please add other packages that you may need BEFORE the SCITEPRESS.sty package.
\usepackage{natbib}
\usepackage{multirow}
\usepackage{SCITEPRESS}
\usepackage{arydshln}
\widowpenalty=10000
\clubpenalty=10000

@@

\cite{Margulieux24} conducted a study on how undergraduate students in
introductory programming courses used generative AI to solve programming
problems in a naturalistic setting. The research focused on examining the
relationship between AI usage and students’ self-regulation strategies,
self-efficacy, and fear of failure in programming. Furthermore, the study
explored how these variables interacted with the characteristics of the learners,
the perceived usefulness of AI, and academic performance. The findings revealed
that students with higher self-efficacy, lower fear of failure, or higher prior
grades tended to use AI less frequently or later in the problem-solving process
and perceived it as less useful compared to their peers. However, no significant
relationship was found between students’ self-regulation strategies and their
use of AI.

\cite{Bastani24} investigated the impact of generative artificial intelligence
(AI), specifically GPT-4, on human learning, with a focus on mathematics
education in a high school. The problem addressed was how the use of generative
AI could affect the acquisition of new skills, which is crucial for long-term
productivity. The method involved a controlled experiment with approximately
one thousand students, who were exposed to two GPT-4-based tutors: a simple
tutor (GPT Base) and another with safeguards designed to promote learning
(GPT Tutor). The results showed that while access to GPT-4 improved performance
on practice exercises (48\% with GPT Base and 127\% with GPT Tutor), the removal
of access to GPT Base led to a 17\% decrease in student performance on exams.
This indicated that the unrestricted use of *GPT Base* could hinder learning.
However, the GPT Tutor was able to mitigate this negative effect.

\cite{Boudouaia24} investigated the impact of ChatGPT-facilitated programming on
college students’ programming behaviors, performance, and perceptions. The
problem addressed was whether the use of ChatGPT in programming education could
influence how students interacted with programming tasks, their learning
outcomes, and their attitudes towards technology. The study employed a
quasi-experimental design involving 82 college students divided into two groups:
one using ChatGPT-assisted programming (CFP) and the other using traditional
self-directed programming (SDP). The method combined mixed data collection
techniques, including behavioral logs, programming performance evaluations, and
interviews. The results showed that students using the CFP mode exhibited more
frequent behaviors such as debugging and reading feedback. Although CFP students
had a higher average score in programming performance, there was no
statistically significant difference compared to the SDP group. However,
students’ perceptions of ChatGPT improved significantly after the intervention,
with increased perceived usefulness, ease of use, and intention to continue
using the tool in the future.

\cite{Sidra24} addressed the growing importance of metacognitive skills in an
AI-enabled workforce. The problem discussed was how AI systems, especially
collaborative tools like Large Language Models (LLMs), impacted human
decision-making and the challenges they posed due to the lack of human-like
metacognition in AI. The method used identified four main characteristics that
differentiated human-AI interaction from human-human collaboration and proposed,
based on the dual-process theory of human thought, that metacognitive thinking
was crucial for the effective use of AI. The results indicated that human
workers needed to develop and apply metacognitive skills, such as planning,
monitoring, and evaluating their thought processes, to mitigate cognitive biases
and improve decision-making when working with AI. The study concluded by
recommending improvements in AI design and training to improve workers'
metacognitive abilities, ensuring more effective collaboration with AI systems.

Learning based on metacognitive strategies has been shown to be effective in
enhancing students’ performance and skills. \cite{Zheng19} investigated the
effects of metacognitive scaffolds on group behavior, performance, and cognitive
load in computer-supported collaborative environments. The results showed
positive impacts on metacognitive behavior and group performance without
increasing the cognitive load. \cite{LiWei23} proposed a metacognition-based
collaborative approach to enhance performance in collaborative programming. The
findings indicated that this approach influenced computational thinking,
critical thinking, and metacognitive awareness. Similarly, \cite{Wang23}
explored how metacognitive instruction affects computational thinking, critical
thinking, and metacognitive skills in collaborative programming, observing
improvements in problem-solving and collaboration. \cite{Khusnul24} investigates
the efficacy of a meta-learning approach in improving metacognitive and creative
skills. This study confirms the effectiveness of metacognition strategies and
elucidates the relationship between meta-learning and metacognition. This
research suggests that metacognition can improve metacognitive abilities,
providing valuable insights into educational technology and course design in
higher education settings.

@@

CharlieBot is an educational chatbot built on ChatGPT 3.5 and enhanced with
Retrieval-Augmented Generation (RAG) technique to improve the precision of
responses provided to students \citep{Sun24}. The RAG approach was developed
using materials from a MOOC, including text, exercises, video transcripts, and
other resources.

@@

The study comprised three phases: data collection, categorization, and analysis.
During the data collection phase, students enrolled in a second-semester Java
course at a public university in Spain were introduced to CharlieBot and given
the freedom to use it without following any prescribed educational methodology.
All data were collected anonymously to ensure that interactions could not be
traced back to individual students or linked to their academic performance.

In the categorization phase, the students' messages were manually classified
into eight distinct categories for analysis, representing different types of
prompt that students use when interacting with the chatbot. Initially,
\cite{Ghimire24} proposed four categories: Debugging Help (DH), Conceptual
Question (CQ), Code Snippet (CS), and Complete Solution (CSO). However, the data
collected indicated the need for additional categories, leading to the inclusion
of four more: Multiple Question (MQ), Student Corrections (SC), Language Change
(LC), and Uncategorized (U). In this study, students were free to use the
chatbot as they wanted, which contributed to the emergence of these additional
categories. Table \ref{tab:categories} presents these categories along with
their respective descriptions. To facilitate the analysis, the prompt topics
were also categorized according to the course subjects, which are: (1) Java, (2)
Object Orientation, (3) Testing, (4) Recursion, (5) Data Structures, and (6)
Sorting and Searching Algorithms.

@@

A total of 625 student messages were categorized in 81 conversations, with
an average of 7.72 messages per conversation. This data offers insights into how
users engage with CharlieBot and the types of prompts they submit. The results
of the analysis are presented in this section, addressing the research questions
outlined in the introduction. Figure \ref{fig:graph1} shows the distribution of
the messages in the categories. The following sections present the results and
discussions of each research question.

@@

Most of the messages are located in the categories Conceptual Question and
Debugging Help corroborate the findings of \cite{Ghimire24} which showed that
the questions are also localized in the same categories.

Multiple Question exercise resolutions represented 14.9\% of the responses.
These prompts typically involved students submitting one or more exercises to
CharlieBot and requesting solutions. The conversation 13 in Figure
\ref{fig:graph2} illustrates a sequence of multiple choice questions.

@@

Figure \ref{fig:graph1} also indicates that approximately 5.8\% of the prompts
written by the students were classified as Uncategorized. These messages include
expressions of gratitude toward the chatbot, such as a simple '\textit{thanks}',
contextual statements like '\textit{I'm reviewing object orientation}', as well
as requests for additional exercises and summaries. Depending on the type of
investigation, there is opportunity for these messages to be further subdivided
into new categories.

@@

Despite some occasional variation (3.54), prompts classified as Uncategorized
exhibit a relatively low average time (1.65) for students to resume interaction
with the chatbot. This variation occurs, in part, due to situations where students
end the conversation with an uncategorized message, such as a simple
\textit{"thanks"} but later choose to resume the interaction.
